---
title: "STAA 566 HW2"
author: "Brian Leonhardi"
date: "2/17/2022"
output: html_document
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding, 
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Code

The purpose of this document is to examine trends in rates of various causes of death over the period 2019-2021 in the U.S., based on quarterly data provided by the CDC.

```{r}
library(dplyr)
library(ggplot2)
library(plotly)

# Read in CDC database on Causes of Death, 2019-2021 by quarter
data <- read.csv("NCHS_-_VSRR_Quarterly_provisional_estimates_for_selected_indicators_of_mortality.csv")

# select Quarter, Time Period, Cause of Death, and Overall Rate
# look at only the crude death rates, filtering out the age-adjusted rates
overall <- data %>% filter(Rate.Type == "Crude") %>%
  select(c("ï..Year.and.Quarter", "Time.Period", "Cause.of.Death", "Overall.Rate"))

# after viewing the rates for various causes, select several causes of interest
# this includes those with the highest rates and those that might be impacted by a pandemic
overall_select <- overall %>% filter(Cause.of.Death %in% c("Heart disease", "Cancer",
        "Coronavirus disease 2019 (COVID-19)", "Falls, ages 65 and over", "Suicide",
        "Homicide", "Drug overdose"))

# split the data by time period: 12 months vs 3 months
overall_select_12 <- overall_select %>% filter(Time.Period == "12 months ending with quarter")
overall_select_3 <- overall_select %>% filter(Time.Period == "3-month period")

```

## Plotting the Results

Each cause of death seems to be plotting on its own y-scale.  As a result, Cancer appears to have a lower rate than drug overdose, while quite the opposite is actually true.

Also, the use of "scale_y_continuous" kept resulting in an error message "Discrete value applied to continuous scale."  

Clearly, much work is needed on this plot.  Eventually I would like to plot the 3-month data and the 12-month data side-by-side for comparison.

```{r}
# plot the the 3-month data over time, color coding by cause of death
overall_select_3_h <- highlight_key(overall_select_3, ~Overall.Rate)
p <- ggplot(overall_select_3_h, aes(ï..Year.and.Quarter, Overall.Rate, group = Cause.of.Death, col = Cause.of.Death))
 p + labs(title = "Rates of Various Causes of Death\n U.S., 2019-2021") +
#         subtitle = "Recorded 1950-2019; Predicted through 2050") +
  geom_point(aes(group = Cause.of.Death, alpha=.5)) + 
  geom_line() +
#   geom_smooth(method = "loess") +
   xlab("Year and Quarter") +
  ylab("Overall Death Rate, Crude") +
   scale_y_discrete(breaks = seq(0, 250, by=50))

```
